---
title: "R Notebook"
output: html_notebook
---

# Chapter 6 Introduction to data visualization

Looking at the numbers and character strings that define a dataset is rarely useful. To convince yourself, print and stare at the US murders data table:
```{r}
library(dslabs)
data(murders)
head(murders)
```

What do you learn from staring at this table? How quickly can you determine which states have the largest populations? Which states have the smallest? How large is a typical state? Is there a relationship between population size and total murders? How do murder rates vary across regions of the country? For most human brains, it is quite difficult to extract this information just by looking at the numbers. In contrast, the answer to all the questions above are readily available from examining this plot:
~

We are reminded of the saying “a picture is worth a thousand words”. Data visualization provides a powerful way to communicate a data-driven finding. In some cases, the visualization is so convincing that no follow-up analysis is required.

The growing availability of informative datasets and software tools has led to increased reliance on data visualizations across many industries, academia, and government. A salient example is news organizations, which are increasingly embracing data journalism and including effective infographics as part of their reporting.

A particularly effective example is a Wall Street Journal article19 showing data related to the impact of vaccines on battling infectious diseases. One of the graphs shows measles cases by US state through the years with a vertical line demonstrating when the vaccine was introduced.

~

Another striking example comes from a New York Times chart20, which summarizes scores from the NYC Regents Exams. As described in the article21, these scores are collected for several reasons, including to determine if a student graduates from high school. In New York City you need a 65 to pass. The distribution of the test scores forces us to notice something somewhat problematic:
~

The most common test score is the minimum passing grade, with very few scores just below the threshold. This unexpected result is consistent with students close to passing having their scores bumped up.

This is an example of how data visualization can lead to discoveries which would otherwise be missed if we simply subjected the data to a battery of data analysis tools or procedures. Data visualization is the strongest tool of what we call exploratory data analysis (EDA). John W. Tukey22, considered the father of EDA, once said,

“The greatest value of a picture is when it forces us to notice what we never expected to see.”

Many widely used data analysis tools were initiated by discoveries made via EDA. EDA is perhaps the most important part of data analysis, yet it is one that is often overlooked.

Data visualization is also now pervasive in philanthropic and educational organizations. In the talks New Insights on Poverty23 and The Best Stats You’ve Ever Seen24, Hans Rosling forces us to notice the unexpected with a series of plots related to world health and economics. In his videos, he uses animated graphs to show us how the world is changing and how old narratives are no longer true.

~

It is also important to note that mistakes, biases, systematic errors and other unexpected problems often lead to data that should be handled with care. Failure to discover these problems can give rise to flawed analyses and false discoveries. As an example, consider that measurement devices sometimes fail and that most data analysis procedures are not designed to detect these. Yet these data analysis procedures will still give you an answer. The fact that it can be difficult or impossible to notice an error just from the reported results makes data visualization particularly important.

In this part of the book, we will learn the basics of data visualization and exploratory data analysis by using three motivating examples. We will use the ggplot2 package to code. To learn the very basics, we will start with a somewhat artificial example: heights reported by students. Then we will cover the two examples mentioned above: 1) world health and economics and 2) infectious disease trends in the United States.

Of course, there is much more to data visualization than what we cover here. The following are references for those who wish to learn more:

-ER Tufte (1983) The visual display of quantitative information. Graphics Press.
-ER Tufte (1990) Envisioning information. Graphics Press.
-ER Tufte (1997) Visual explanations. Graphics Press.
-WS Cleveland (1993) Visualizing data. Hobart Press.
-WS Cleveland (1994) The elements of graphing data. CRC Press.
-A Gelman, C Pasarica, R Dodhia (2002) Let’s practice what we preach: Turning tables into graphs. The American Statistician 56:121-130.
-NB Robbins (2004) Creating more effective graphs. Wiley.
-A Cairo (2013) The functional art: An introduction to information graphics and visualization. New Riders.
-N Yau (2013) Data points: Visualization that means something. Wiley.

We also do not cover interactive graphics, a topic that is too advanced for this book. Some useful resources for those interested in learning more can be found below:

https://shiny.rstudio.com/
https://d3js.org/

# Chapter 7 ggplot2

Exploratory data visualization is perhaps the greatest strength of R. One can quickly go from idea to data to plot with a unique balance of flexibility and ease. For example, Excel may be easier than R for some plots, but it is nowhere near as flexible. D3.js may be more flexible and powerful than R, but it takes much longer to generate a plot.

Throughout the book, we will be creating plots using the ggplot225 package.
```{r}
library(dplyr)
library(ggplot2)
```

Many other approaches are available for creating plots in R. In fact, the plotting capabilities that come with a basic installation of R are already quite powerful. There are also other packages for creating graphics such as grid and lattice. We chose to use ggplot2 in this book because it breaks plots into components in a way that permits beginners to create relatively complex and aesthetically pleasing plots using syntax that is intuitive and comparatively easy to remember.

One reason ggplot2 is generally more intuitive for beginners is that it uses a grammar of graphics26, the gg in ggplot2. This is analogous to the way learning grammar can help a beginner construct hundreds of different sentences by learning just a handful of verbs, nouns and adjectives without having to memorize each specific sentence. Similarly, by learning a handful of ggplot2 building blocks and its grammar, you will be able to create hundreds of different plots.

Another reason ggplot2 is easy for beginners is that its default behavior is carefully chosen to satisfy the great majority of cases and is visually pleasing. As a result, it is possible to create informative and elegant graphs with relatively simple and readable code.

One limitation is that ggplot2 is designed to work exclusively with data tables in tidy format (where rows are observations and columns are variables). However, a substantial percentage of datasets that beginners work with are in, or can be converted into, this format. An advantage of this approach is that, assuming that our data is tidy, ggplot2 simplifies plotting code and the learning of grammar for a variety of plots.

To use ggplot2 you will have to learn several functions and arguments. These are hard to memorize, so we highly recommend you have the ggplot2 cheat sheet handy. You can get a copy here: https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf or simply perform an internet search for “ggplot2 cheat sheet”.

## 7.1 The components of a graph

We will construct a graph that summarizes the US murders dataset that looks like this:
~

We can clearly see how much states vary across population size and the total number of murders. Not surprisingly, we also see a clear relationship between murder totals and population size. A state falling on the dashed grey line has the same murder rate as the US average. The four geographic regions are denoted with color, which depicts how most southern states have murder rates above the average.

This data visualization shows us pretty much all the information in the data table. The code needed to make this plot is relatively simple. We will learn to create the plot part by part.

The first step in learning ggplot2 is to be able to break a graph apart into components. Let’s break down the plot above and introduce some of the ggplot2 terminology. The main three components to note are:

-Data: The US murders data table is being summarized. We refer to this as the data component.
-Geometry: The plot above is a scatterplot. This is referred to as the geometry component. Other possible geometries are barplot, histogram, smooth densities, qqplot, and boxplot. We will learn more about these in the Data Visualization part of the book.
-Aesthetic mapping: The plot uses several visual cues to represent the information provided by the dataset. The two most important cues in this plot are the point positions on the x-axis and y-axis, which represent population size and the total number of murders, respectively. Each point represents a different observation, and we map data about these observations to visual cues like x- and y-scale. Color is another visual cue that we map to region. We refer to this as the aesthetic mapping component. How we define the mapping depends on what geometry we are using.

We also note that:

-The points are labeled with the state abbreviations.
-The range of the x-axis and y-axis appears to be defined by the range of the data. They are both on log-scales.
-There are labels, a title, a legend, and we use the style of The Economist magazine.
We will now construct the plot piece by piece.

We start by loading the dataset:
```{r}
library(dslabs)
data(murders)
```

## 7.2 ```ggplot``` objects
The first step in creating a ggplot2 graph is to define a ```ggplot``` object. We do this with the function ```ggplot```, which initializes the graph. If we read the help file for this function, we see that the first argument is used to specify what data is associated with this object:
```{r}
ggplot(data = murders)
```

We can also pipe the data in as the first argument. So this line of code is equivalent to the one above:
```{r}
murders %>% ggplot()
```

~

It renders a plot, in this case a blank slate since no geometry has been defined. The only style choice we see is a grey background.

What has happened above is that the object was created and, because it was not assigned, it was automatically evaluated. But we can assign our plot to an object, for example like this:
```{r}
p <- ggplot(data = murders)
class(p)
```

To render the plot associated with this object, we simply print the object p. The following two lines of code each produce the same plot we see above:
```{r}
print(p)
p
```

## 7.3 Geometries

In ```ggplot2``` we create graphs by adding layers. Layers can define geometries, compute summary statistics, define what scales to use, or even change styles. To add layers, we use the symbol ```+```. In general, a line of code will look like this:

DATA %>% ggplot() + LAYER 1 + LAYER 2 + … + LAYER N

Usually, the first added layer defines the geometry. We want to make a scatterplot. What geometry do we use?

Taking a quick look at the cheat sheet, we see that the function used to create plots with this geometry is ```geom_point```.

Geometry function names follow the pattern: ```geom_X``` where X is the name of the geometry. Some examples include ```geom_point```, ```geom_bar```, and ```geom_histogram```.

For ```geom_point``` to run properly we need to provide data and a mapping. We have already connected the object ```p``` with the ```murders``` data table, and if we add the layer ```geom_point``` it defaults to using this data. To find out what mappings are expected, we read the Aesthetics section of the help file ```geom_point``` help file:
```{r, eval=F}
> Aesthetics
> 
> geom_point understands the following aesthetics (required aesthetics are in bold):
>
> x
>
> y
> 
> alpha
>
> colour
```

and, as expected, we see that at least two arguments are required ```x``` and ```y```.

## 7.4 Aesthetic mappings

Aesthetic mappings describe how properties of the data connect with features of the graph, such as distance along an axis, size, or color. The ```aes``` function connects data with what we see on the graph by defining aesthetic mappings and will be one of the functions you use most often when plotting. The outcome of the ```aes``` function is often used as the argument of a geometry function. This example produces a scatterplot of total murders versus population in millions:
```{r}
murders %>% ggplot() + 
  geom_point(aes(x = population/10^6, y = total))
```

We can drop the ```x =``` and ```y =``` if we wanted to since these are the first and second expected arguments, as seen in the help page.

Instead of defining our plot from scratch, we can also add a layer to the ```p``` object that was defined above as ```p <- ggplot(data = murders)```:
```{r}
p + geom_point(aes(population/10^6, total))
```

~

The scale and labels are defined by default when adding this layer. Like dplyr functions, ```aes``` also uses the variable names from the object component: we can use ```population``` and ```total``` without having to call them as ```murders$population``` and ```murders$total```. The behavior of recognizing the variables from the data component is quite specific to ```aes```. With most functions, if you try to access the values of ```population``` or ```total``` outside of ```aes``` you receive an error.

## 7.5 Layers

A second layer in the plot we wish to make involves adding a label to each point to identify the state. The ```geom_label``` and ```geom_text``` functions permit us to add text to the plot with and without a rectangle behind the text, respectively.

Because each point (each state in this case) has a label, we need an aesthetic mapping to make the connection between points and labels. By reading the help file, we learn that we supply the mapping between point and label through the ```label``` argument of ```aes```. So the code looks like this:
```{r}
p + geom_point(aes(population/10^6, total)) +
  geom_text(aes(population/10^6, total, label = abb))
```

~
We have successfully added a second layer to the plot.

As an example of the unique behavior of aes mentioned above, note that this call:
```{r}
p_test <- p + geom_text(aes(population/10^6, total, label = abb))
```
is fine, whereas this call:
```{r, eval=F}
p_test <- p + geom_text(aes(population/10^6, total), label = abb)
```
will give you an error since ```abb``` is not found because it is outside of the ```aes``` function. The layer ```geom_text``` does not know where to find ```abb``` since it is a column name and not a global variable.

### 7.5.1 Tinkering with arguments

Each geometry function has many arguments other than ```aes``` and ```data```. They tend to be specific to the function. For example, in the plot we wish to make, the points are larger than the default size. In the help file we see that ```size``` is an aesthetic and we can change it like this:
```{r}
p + geom_point(aes(population/10^6, total), size = 3) +
  geom_text(aes(population/10^6, total, label = abb))
```
~

```size``` is not a mapping: whereas mappings use data from specific observations and need to be inside ```aes()```, operations we want to affect all the points the same way do not need to be included inside ```aes```.

Now because the points are larger it is hard to see the labels. If we read the help file for ```geom_text```, we see the ```nudge_x``` argument, which moves the text slightly to the right or to the left:
```{r}
p + geom_point(aes(population/10^6, total), size = 3) +
  geom_text(aes(population/10^6, total, label = abb), nudge_x = 1.5)
```

~

This is preferred as it makes it easier to read the text. In Section 7.11 we learn a better way of assuring we can see the points and the labels.

## 7.6 Global versus local aesthetic mappings
In the previous line of code, we define the mapping ```aes(population/10^6, total)``` twice, once in each geometry. We can avoid this by using a global aesthetic mapping. We can do this when we define the blank slate ```ggplot``` object. Remember that the function ```ggplot``` contains an argument that permits us to define aesthetic mappings:
```{r}
args(ggplot)
```

If we define a mapping in ```ggplot```, all the geometries that are added as layers will default to this mapping. We redefine ```p```:
```{r}
p <- murders %>% ggplot(aes(population/10^6, total, label = abb))
```
and then we can simply write the following code to produce the previous plot:
```{r}
p + geom_point(size = 3) + 
  geom_text(nudge_x = 1.5)
```

We keep the ```size``` and ```nudge_x``` arguments in ```geom_point``` and ```geom_text```, respectively, because we want to only increase the size of points and only nudge the labels. If we put those arguments in ```aes``` then they would apply to both plots. Also note that the ```geom_point``` function does not need a ```label``` argument and therefore ignores that aesthetic.

If necessary, we can override the global mapping by defining a new mapping within each layer. These local definitions override the global. Here is an example:
```{r}
p + geom_point(size = 3) +  
  geom_text(aes(x = 10, y = 800, label = "Hello there!"))
```

~
Clearly, the second call to ```geom_text``` does not use ```population``` and ```total```.

## 7.7 Scales

First, our desired scales are in log-scale. This is not the default, so this change needs to be added through a scales layer. A quick look at the cheat sheet reveals the ```scale_x_continuous``` function lets us control the behavior of scales. We use them like this:
```{r}
p + geom_point(size = 3) +  
  geom_text(nudge_x = 0.05) + 
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10") 
```

~

Because we are in the log-scale now, the nudge must be made smaller.

This particular transformation is so common that ggplot2 provides the specialized functions ```scale_x_log10``` and ```scale_y_log10```, which we can use to rewrite the code like this:
```{r}
p + geom_point(size = 3) +  
  geom_text(nudge_x = 0.05) + 
  scale_x_log10() +
  scale_y_log10() 
```

## 7.8 Labels and titles

Similarly, the cheat sheet quickly reveals that to change labels and add a title, we use the following functions:
```{r}
p + geom_point(size = 3) +  
  geom_text(nudge_x = 0.05) + 
  scale_x_log10() +
  scale_y_log10() +
  xlab("Populations in millions (log scale)") + 
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010")
```
~
We are almost there! All we have left to do is add color, a legend, and optional changes to the style.